Supplement to ``{\mytitle}''

### 2024-02-16 si/calibrateMod3.Rnw
Finished addressing reviewer comments

# Get repeated estimates of unit log-likelihoods
unitLikesM3 <- best_mod %>%
  pivot_longer(
    cols = -which,
    names_to = "unit",
    names_prefix = "V",
    values_to = "logLik"
  ) %>%
  select(-which) %>%
  group_by(unit) %>%
  summarize(
    ll_est = pomp::logmeanexp(logLik),
    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
  ) %>%
  ungroup() %>%
  mutate(unit = as.numeric(unit)) %>%
  arrange(unit)

@@

# Combine model likelihood table to benchmarks
bench_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  select(-se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))

bench_diffs <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  rename(ll_mod3 = est, ll_bench = ll) %>%
  mutate(unit = unlist(myLabeller(unit)),
         ll_diff = ll_mod3 - ll_bench) %>%
  arrange(ll_diff)

@@

We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.

@@

# Set the model parameters to the best results from previous search
coef(h3)[names(best_params)] <- best_params

@@

@@

\begin{figure}[!ht]
<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
  geom_point(aes(col = type)) +
  ylab("Unit") +
  xlab("Log Likelihood") +
  scale_color_manual(
    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
  ) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank())
@
\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to Hurricane Matthew.}
\end{figure}

The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for units \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.

@@

Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e. $f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.

@@

    # Get "noVacc" scenario parameters
    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
    h3Spat <- h3
    p3 <- best_params
    coef(h3Spat) <- p3
    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par

@@

    # Using parameters and end state values, obtain projections
    h3_V0_sims <- project_from_filter(
      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
      nsims = 1000, seed = 184697
    )

>>>>

unitLikesM3_final <- readRDS("../model3/run_level_3/mod3_eval_res.rds")[['u_cond_ll']] %>%
  t() %>% 
  as_tibble() %>% 
  rename(ll_est = est, ll_se = se) %>% 
  mutate(unit = 1:10) %>% 
  select(unit, ll_est, ll_se)

# Get repeated estimates of unit log-likelihoods
# unitLikesM3 <- best_mod %>%
#   pivot_longer(
#     cols = -which,
#     names_to = "unit",
#     names_prefix = "V",
#     values_to = "logLik"
#   ) %>%
#   select(-which) %>%
#   group_by(unit) %>%
#   summarize(
#     ll_est = pomp::logmeanexp(logLik),
#     ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
#   ) %>%
#   ungroup() %>%
#   mutate(unit = as.numeric(unit)) %>%
#   arrange(unit)

@@

# Combine model likelihood table to benchmarks
bench_M3_joined_final <- dplyr::left_join(
  x = unitLikesM3_final,
  y = unitLikes_bench,
  by = "unit"
) %>%
  select(-ll_se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))

bench_diffs_final <- dplyr::left_join(
  x = unitLikesM3_final,
  y = unitLikes_bench,
  by = "unit"
) %>%
  rename(ll_mod3 = ll_est, ll_bench = ll) %>%
  mutate(unit = unlist(myLabeller(unit)),
         ll_diff = ll_mod3 - ll_bench) %>%
  arrange(ll_diff)

@@

We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
The resulting model has a log-likelihood value of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'], 1)}$.
The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'] - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.

@@

@@

    # Set the model parameters to the best results from previous search
    coef(h3)[names(h3_final_mod_pars)] <- h3_final_mod_pars

@@

\begin{figure}[!ht]
<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
ggplot(bench_M3_joined_final, aes(x = ll, y = reorder(unit, ll))) +
  geom_point(aes(col = type)) +
  ylab("Unit") +
  xlab("Log Likelihood") +
  scale_color_manual(
    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
  ) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank())
@
\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to Hurricane Matthew.}
\end{figure}

The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for the departments \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.

@@

Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e. $f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data. 
This, coupled with the fact that the model fails to beat a simple statistical benchmark on data on the four departments that experienced the majority of observed cholera cases, suggests that one should be careful when making inference using this model.

@@

    # Get "noVacc" scenario parameters
    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
    h3Spat <- h3
    p3 <- h3_final_mod_pars

    coef(h3Spat) <- p3
    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par

@@

    # Using parameters and end state values, obtain projections
    h3_V0_sims <- project_from_filter(
      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
      nsims = 500, seed = 184697
    )

### 2024-02-16 si/initialValues.Rnw
Finished addressing reviewer comments

The AIC of this alternative approach is $\Sexpr{myround(mod2_newInit_AIC, 1)}$, compared to the fixed approach with an AIC of $\Sexpr{myround(mod2_AIC, 1)}$. 
This alternative initialization approach results in quantitative improvement to the model-fit, but does not result in qualitative differences in the conclusions made using this model. Figure~\ref{fig:h2InitialTraj} displays the trajectory of the model with this alternate initialization model. 
Table~\ref{tab:InitValues} gives the estimated initial values. 
The estimated value of the latent state is similar to the fixed value in Artibonite and Centre, where the largest number of cases are present at the start of the epidemic. 
Because of this, the qualitative dynamics do not differ by much when the parameters are estimated versus held constant, despite the improvement in model fit, measured by AIC.

@@

\begin{figure}[!ht]
<<h2InitialTraj, echo=FALSE, message=FALSE, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:h2InitialTraj}
{\figTitle Simulated trajectory of alternate initialization of Model~2.}
The black line shows the nationally aggregated weekly cholera incidence data.
The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
Compare to Fig.~4 of the main article.}
\end{figure}

>>>>

The AIC of this alternative approach is $\Sexpr{myround(mod2_newInit_AIC, 1)}$, compared to the fixed approach with an AIC of $\Sexpr{myround(mod2_AIC, 1)}$. 
This alternative initialization approach results in quantitative improvement to the model-fit, but does not result in qualitative differences in the conclusions made using this model. Figure~\ref{fig:h2InitialTraj} displays the trajectory of the model with this alternate initialization model. 
Table~\ref{tab:InitValues} gives the estimated initial values. 
The estimated value of the latent state is similar to the fixed value in Artibonite and Centre, where the largest number of cases are present at the start of the epidemic. 
Because of this, the qualitative dynamics do not differ by much when the parameters are estimated versus held constant, despite the improvement in model fit measured by AIC.

@@

\begin{figure}[!ht]
<<h2InitialTraj, echo=FALSE, message=FALSE, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:h2InitialTraj}
{\figTitle Simulated trajectory of alternate initialization of Model~2.}
The black line shows the nationally aggregated weekly cholera incidence data.
The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
Compare to Fig.~4 of the article.}
\end{figure}
